LCS Tool: A Computational Platform for Lagrangian Coherent Structures

### 2014-12-12 main.tex
Add FIXME comments

\begin{abstract}
We give an algorithmic introduction to Lagrangian coherent structures (LCSs) using a newly developed computational engine, LCS Tool.
LCSs are most repelling, attracting and shearing material lines that form the centrepieces of observed tracer patterns in two-dimensional unsteady dynamical systems.
LCS Tool implements the latest geodesic theory for LCSs in two-dimensional flows, uncovering key transport barriers in unsteady flow velocity data as explicit solutions of differential equations.
LCS Tool makes theoretical results accessible to the fluid mechanics community since implementing these results directly could be time consuming.
After a review of the underlying theory, we explain the numerical algorithms used by LCS Tool, and illustrate its capabilities on three unsteady fluid flow examples: a double gyre, a Bickley jet and an ocean dataset for the Agulhas current.
\end{abstract}

@@

Lagrangian Coherent Structures (LCSs) are evolving organizing centres of trajectory patterns in non-autonomous dynamical systems\parencite{haller00:_lagran,peacock13:_lagran,haller15:_langr_coher_struc}. Applications of LCSs include oceanic and atmospheric flows\parencite{beron-vera13:_objec_agulh,koh02:_hyper}, biological transport problems\parencite{wilson09:_lagran_reynol,tallapragada11:_lagran,huhn12:_south_indian_ocean_count_madag}, aeronautics\parencite{tang10:_accur_lagran_hong_kong_inter_airpor}, celestial mechanics\parencite{gawlik09:_lagran}, crowd dynamics\parencite{ali07:_lagran_partic_dynam_approac_crowd}, and aperiodically forced mechanical oscillators\parencite{hadjighasem13:_detec_kam}.

@@

\subsection{Elliptic LCSs}
\label{sec:Elliptic LCSs}

@@

\subsection{Hyperbolic LCSs}
\label{sec:Hyperbolic LCSs}

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Resolution & Runtime (s) & Memory (MB)\tabularnewline
\hline \hline
$500 \times 251$ & 29 & 55\tabularnewline
\hline
$750 \times 376$ & 63 & 123\tabularnewline
\hline
$1000 \times 501$ & 112 & 223\tabularnewline
\hline
\end{tabular}
\end{center}
\caption{
Runtime and memory use associated with vector integration at different resolutions.
Results are from the function \lstinline!eig_cgStrain! applied to the double gyre presented in \cref{sec:double gyre}.
Results obtained with an Intel Core i3-4130 processor.
}
\label{t:vector integration memory}
\end{table}

@@

Velocity fields defined by datasets require preprocessing before they are used in the numerical integration of \cref{eq:velocity}. This requires spatial and temporal interpolation that enables the evaluation of the velocity function at arbitrary points in $U$ and at arbitrary times between $t_-$ and $t_+$. In \cref{sec:oceandataset}, we present an ocean dataset example with details of possible interpolation functions.

@@

As discussed in \cref{sec:Elliptic LCSs}, positions of elliptic LCSs at time $t_0$ are found as closed orbits of the $\eta_\pm^\lambda$ vector fields defined in \cref{eq:eta}. We find such orbits by integrating \cref{eq:etafields} from points of an appropriately chosen section (Poincare section), and evaluating the first return map (Poincare map) onto this section. A $\lambda$-line returning to its starting point is then an elliptic LCS. The outermost member of a family of closed $\lambda$-lines (obtained by varying $\lambda$) is a coherent Lagrangian vortex boundary\parencite{haller13:_coher_lagran,haller14:_adden_coher_lagran}.

@@

As discussed in \cref{sec:Hyperbolic LCSs}, hyperbolic LCS positions at time $t_0$ are found as the strongest repelling curves of the vector field \cref{eq:shrink line} (repelling LCSs), and strongest attracting curves of the vector field \cref{eq:stretch line} (attracting LCSs).
By repulsion and attraction we mean a property of the LCS (as an evolving material line) under the flow map $F_{t_0}^t$.
We identify the strongest repelling shrink lines as those crossing a local maximum of the $\lambda_2(x_0)$ field.
Similarly, we identify the strongest attracting stretch lines as those crossing a local minimum of the $\lambda_1(x_0)$ field.
These local maxima and minima of the appropriate $\lambda_i(x_0)$ eigenvalue field can be thought of as the extensions of the concept of saddle points to the present finite-time, temporally aperiodic flow setting.

@@

\subsection{Double gyre}
\label{sec:double gyre}

@@

\lstinputlisting[caption={LCS Tool commands for double gyre elliptic LCSs. Abridged subset of the LCS Tool script \lstinline!demo/double_gyre/elliptic_hyperbolic_lcs.m!},
                 label=l:double gyre elliptic LCS]{listings/double_gyre/elliptic_lcs_subset.m}

@@

The final example demonstrates the use of LCS Tool on velocity data derived from satellite-observed sea-surface heights under the geostrophic approximation. In contrast to the previous two analytic examples, the velocity field is available only with discrete temporal and spatial resolution. Our region of interest is a small domain in the South Atlantic Ocean, where exceptionally coherent eddies, Agulhas rings, were recently found by \textcite{haller13:_coher_lagran,haller14:_adden_coher_lagran} using the theory we surveyed in \cref{sec:Elliptic LCSs}.

>>>>

\begin{abstract}
We give an algorithmic introduction to Lagrangian coherent structures (LCSs) using a newly developed computational engine, LCS Tool.
% FIXME "centrepiece" is vague and may confuse with centre fixed point in classical dynamical systems
LCSs are most repelling, attracting and shearing material lines that form the centrepieces of observed tracer patterns in two-dimensional unsteady dynamical systems.
LCS Tool implements the latest geodesic theory for LCSs in two-dimensional flows, uncovering key transport barriers in unsteady flow velocity data as explicit solutions of differential equations.
LCS Tool makes theoretical results accessible to the fluid mechanics community since implementing these results directly could be time consuming.
After a review of the underlying theory, we explain the numerical algorithms used by LCS Tool, and illustrate its capabilities on three unsteady fluid flow examples: a double gyre, a Bickley jet and an ocean dataset for the Agulhas current.
\end{abstract}

@@

% FIXME "Centres" may be associated with center fixed points in the context of classical dynamical systems
Lagrangian Coherent Structures (LCSs) are evolving organizing centres of trajectory patterns in non-autonomous dynamical systems\parencite{haller00:_lagran,peacock13:_lagran,haller15:_langr_coher_struc}. Applications of LCSs include oceanic and atmospheric flows\parencite{beron-vera13:_objec_agulh,koh02:_hyper}, biological transport problems\parencite{wilson09:_lagran_reynol,tallapragada11:_lagran,huhn12:_south_indian_ocean_count_madag}, aeronautics\parencite{tang10:_accur_lagran_hong_kong_inter_airpor}, celestial mechanics\parencite{gawlik09:_lagran}, crowd dynamics\parencite{ali07:_lagran_partic_dynam_approac_crowd}, and aperiodically forced mechanical oscillators\parencite{hadjighasem13:_detec_kam}.

@@

\subsection{Elliptic LCSs}
\label{s:Elliptic LCSs}

@@

\subsection{Hyperbolic LCSs}
\label{s:Hyperbolic LCSs}

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Resolution & Runtime (s) & Memory (MB)\tabularnewline
\hline \hline
$500 \times 251$ & 29 & 55\tabularnewline
\hline
$750 \times 376$ & 63 & 123\tabularnewline
\hline
$1000 \times 501$ & 112 & 223\tabularnewline
\hline
\end{tabular}
\end{center}
\caption{
Runtime and memory use associated with vector integration at different resolutions.
Results are from the function \lstinline!eig_cgStrain! applied to the double gyre presented in \cref{s:double gyre}.
Results obtained with an Intel Core i3-4130 processor.
}
\label{t:vector integration memory}
\end{table}

@@

Velocity fields defined by datasets require preprocessing before they are used in the numerical integration of \cref{eq:velocity}. This requires spatial and temporal interpolation that enables the evaluation of the velocity function at arbitrary points in $U$ and at arbitrary times between $t_-$ and $t_+$. In \cref{s:oceandataset}, we present an ocean dataset example with details of possible interpolation functions.

@@

As discussed in \cref{s:Elliptic LCSs}, positions of elliptic LCSs at time $t_0$ are found as closed orbits of the $\eta_\pm^\lambda$ vector fields defined in \cref{eq:eta}. We find such orbits by integrating \cref{eq:etafields} from points of an appropriately chosen section (Poincare section), and evaluating the first return map (Poincare map) onto this section. A $\lambda$-line returning to its starting point is then an elliptic LCS. The outermost member of a family of closed $\lambda$-lines (obtained by varying $\lambda$) is a coherent Lagrangian vortex boundary\parencite{haller13:_coher_lagran,haller14:_adden_coher_lagran}.

@@

As discussed in \cref{s:Hyperbolic LCSs}, hyperbolic LCS positions at time $t_0$ are found as the strongest repelling curves of the vector field \cref{eq:shrink line} (repelling LCSs), and strongest attracting curves of the vector field \cref{eq:stretch line} (attracting LCSs).
By repulsion and attraction we mean a property of the LCS (as an evolving material line) under the flow map $F_{t_0}^t$.
We identify the strongest repelling shrink lines as those crossing a local maximum of the $\lambda_2(x_0)$ field.
Similarly, we identify the strongest attracting stretch lines as those crossing a local minimum of the $\lambda_1(x_0)$ field.
These local maxima and minima of the appropriate $\lambda_i(x_0)$ eigenvalue field can be thought of as the extensions of the concept of saddle points to the present finite-time, temporally aperiodic flow setting.

@@

\subsection{Double gyre}
\label{s:double gyre}

@@

\lstinputlisting[
caption={LCS Tool commands for double gyre elliptic LCSs.
Abridged subset of the LCS Tool script \lstinline!demo/double_gyre/elliptic_hyperbolic_lcs.m!},
label=l:double gyre elliptic LCS]
{listings/double_gyre/elliptic_lcs_subset.m}

@@

The final example demonstrates the use of LCS Tool on velocity data derived from satellite-observed sea-surface heights under the geostrophic approximation. In contrast to the previous two analytic examples, the velocity field is available only with discrete temporal and spatial resolution. Our region of interest is a small domain in the South Atlantic Ocean, where exceptionally coherent eddies, Agulhas rings, were recently found by \textcite{haller13:_coher_lagran,haller14:_adden_coher_lagran} using the theory we surveyed in \cref{s:Elliptic LCSs}.
